---
title: "PHTS EDA"
author: "R. Jerome Dixon"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
format:
  html:
    toc: true
    toc-depth: 5
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
    default-image-extension: svg
    dpi: 600
---

## Overview

This exploratory data analysis examines pediatric heart transplant data using variables identified in the Wisotzkey et al paper. The analysis includes data preprocessing, model development, and evaluation of prediction performance.

```{r echo=FALSE, warning=FALSE, message=FALSE}

library(here)
library(readr)
library(haven)
library(purrr)
library(dplyr)
library(glmnet)
library(caret)
library(yardstick)
library(janitor)
library(lubridate)

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

wisotzkey_variables <- read_csv(here("data","wisotzkey_variables.csv"), show_col_types = FALSE)

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Set directory
data_dir <- here("data")

# List SAS files
sas_files <- list.files(data_dir, pattern = "\\.sas7bdat$", full.names = TRUE)

# Named list: file base names (without extension) as names
sas_data <- sas_files %>%
  set_names(~ tools::file_path_sans_ext(basename(.))) %>%
  map(read_sas)

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

mcsd <- sas_data$mcsd %>% janitor::clean_names()

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

eval <- sas_data$pretxeval %>% janitor::clean_names()

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

tx <- sas_data$transplant %>% janitor::clean_names()

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

followups <- sas_data$posttxplfollowup %>% janitor::clean_names()

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

wisotzkey_vars <- wisotzkey_variables[[1]] %>% as.character()
wisotzkey_clean <- tolower(gsub("[^a-zA-Z0-9]+", "_", wisotzkey_vars))
wisotzkey_clean

```

```{r echo=FALSE}

# Calculate BMI, eGFR, and Listing Year
tx <- tx %>%
  mutate(
    age_listing = as.double(age_listing),
    age_txpl = as.double(age_txpl),
    bmi_txpl = (weight_txpl / (height_txpl^2)) * 703,
    egfr_tx = 0.413 * height_txpl / txcreat_r,
    listing_year = as.integer(floor(txpl_year - (age_txpl - age_listing)))
  )

```

```{r}
tx %>% colnames()
```

```{r echo=FALSE}

wisotzkey_name_map <- c(
  "primary_etiology"               = "prim_dx",
  "mcsd_at_transplant"             = "txmcsd",
  "single_ventricle_chd"           = "chd_sv",
  "surgeries_prior_to_listing"     = "hxsurg",
  "serum_albumin_at_transplant"    = "txsa_r",
  "bun_at_transplant"              = "txbun_r",
  "ecmo_at_transplant"             = "txecmo",
  "transplant_year"                = "txpl_year",
  "recipient_weight_at_transplant" = "weight_txpl",
  "alt_at_transplant"              = "txalt",
  "bmi_at_transplant"              = "bmi_txpl",     # computed
  "pra_max_at_listing"             = "lsfprat",      # fallback: use lsfprab if missing
  "egfr_at_transplant"             = "egfr_tx",      # computed from height/creatinine
  "medical_history_at_listing"     = "hxmed",
  "listing_year"                   = "listing_year"  # computed from txpl_year, age_listing, age_txpl
)

```

```{r echo=FALSE}

tx <- tx %>%
  mutate(
    ev_time = pmin(int_dead, int_graft_loss, na.rm = TRUE),
    ev_type = pmax(dtx_patient, graft_loss, na.rm = TRUE),
    outcome = case_when(
      ev_type == 1 & ev_time < 1 ~ 1L,
      ev_type == 0 & ev_time < 1 ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  filter(!is.na(outcome))

# Impute values per Wisotzkey paper
tx_model <- tx %>%
  mutate(
    prim_dx = ifelse(is.na(prim_dx), "cardiomyopathy", prim_dx),
    txmcsd = ifelse(is.na(txmcsd), 0L, txmcsd),
    chd_sv = ifelse(is.na(chd_sv), 0L, chd_sv),
    hxsurg = ifelse(is.na(hxsurg), 0L, hxsurg),
    txsa_r = ifelse(is.na(txsa_r), 3.7, txsa_r),
    txbun_r = ifelse(is.na(txbun_r), 16, txbun_r),
    txecmo = ifelse(is.na(txecmo), 0L, txecmo),
    txpl_year = ifelse(is.na(txpl_year), 2015, txpl_year),
    weight_txpl = ifelse(is.na(weight_txpl), 35, weight_txpl),
    txalt = ifelse(is.na(txalt), 29, txalt),
    bmi_txpl = ifelse(is.na(bmi_txpl), 17, bmi_txpl),
    lsfprat = ifelse(is.na(lsfprat), 0, lsfprat),
    egfr_tx = ifelse(is.na(egfr_tx), 97, egfr_tx),
    hxmed = ifelse(is.na(hxmed), 1L, hxmed),
    listing_year = ifelse(is.na(listing_year), 2014, listing_year)
  ) %>%
  rename(!!!setNames(wisotzkey_name_map, names(wisotzkey_name_map))) %>%
  select(ptid_e, all_of(names(wisotzkey_name_map)), outcome) 

```

```{r echo=FALSE}

tx_model <- tx_model %>%
  mutate(
    serum_albumin_at_transplant_bin = as.integer(serum_albumin_at_transplant > 3.7),
    pra_max_at_listing_bin          = as.integer(pra_max_at_listing > 0),
    alt_at_transplant_bin           = as.integer(alt_at_transplant > 29),
    single_ventricle_chd_bin        = as.integer(single_ventricle_chd > 0),
    egfr_at_transplant_bin          = as.integer(egfr_at_transplant > 97),
    bun_at_transplant_bin           = as.integer(bun_at_transplant > 16),
    bmi_at_transplant_bin           = as.integer(bmi_at_transplant > 17)
  ) %>%
  relocate(
    serum_albumin_at_transplant_bin,
    pra_max_at_listing_bin,
    alt_at_transplant_bin,
    single_ventricle_chd_bin,
    egfr_at_transplant_bin,
    bun_at_transplant_bin,
    bmi_at_transplant_bin,
    .after = medical_history_at_listing
  )

```

### LASSO Model

```{r warning=FALSE}

# Split into training (2010–2013) and test (2014)
train <- tx_model %>% filter(transplant_year >= 2010 & transplant_year <= 2013)
test  <- tx_model %>% filter(transplant_year == 2014)

# Save patient IDs separately
ptid_e_train <- train$ptid_e
ptid_e_test <- test$ptid_e

# Remove ptid_e from feature set before model.matrix()
x_train <- model.matrix(outcome ~ . - ptid_e - 1, data = train)
x_test <- model.matrix(outcome ~ . - ptid_e - 1, data = test)

y_train <- train$outcome
y_test <- test$outcome

# Fit LASSO model
cvfit <- cv.glmnet(x_train, y_train, family = "binomial", alpha = 1)

# Predict and evaluate
pred_probs <- predict(cvfit, newx = x_test, s = "lambda.min", type = "response")

# Extract coefficients at lambda.min
important_vars <- coef(cvfit, s = "lambda.min")

# Create test results and misclassified data frames
test_results <- test %>%
  mutate(
    ptid_e = ptid_e_test,
    pred_prob = as.vector(pred_probs),  # ensure numeric vector
    predicted = as.vector(ifelse(pred_probs >= 0.5, 1, 0)),  # flatten matrix result
    correct = ifelse(predicted == outcome, "Correct", "Misclassified")
  )

misclassified <- test_results %>%
  filter(predicted != outcome)

important_vars

```

### CART Model

```{r}

# Create CART model for misclassified cases
library(rpart)
if (nrow(misclassified) > 10) {
  # Create dummy target
  misclassified$fake_y <- 1:nrow(misclassified)
  
  # Train tree
  tree_misclassified <- rpart(
    fake_y ~ . - ptid_e - predicted - pred_prob - outcome - fake_y, 
    data = misclassified, 
    method = "anova",
    control = rpart.control(minsplit = 5, cp = 0.01)
  )
  
  # Extract variable importance
  var_importance <- tree_misclassified$variable.importance
  
  if (!is.null(var_importance)) {
    var_importance_df <- data.frame(
      variable = names(var_importance),
      importance = as.numeric(var_importance)
    ) %>%
    arrange(desc(importance))
    
    # Save
    write.csv(var_importance_df, "reports/misclassified1_variable_importance.csv", row.names = FALSE)
  } else {
    cat("No variable importance extracted.\n")
  }
} else {
  cat("Not enough samples to analyze.\n")
}

var_importance_df <- tibble(
  variable = names(var_importance),
  importance = as.numeric(var_importance)
) %>%
  mutate(importance = importance / max(importance))  # Normalize to [0,1]


var_importance_df
```

### LASSO and CART Feature Importances

```{r}

lasso_vars <- rownames(important_vars)[important_vars[,1] != 0]
cart_vars <- var_importance_df$variable

# Intersection
intersect_vars <- intersect(lasso_vars, cart_vars)
cat("Features shared between LASSO and CART explanations:\n")
print(intersect_vars)

# LASSO-only
cat("\nLASSO-only features:\n")
print(setdiff(lasso_vars, cart_vars))

# CART-only
cat("\nCART-only features (missed by LASSO):\n")
print(setdiff(cart_vars, lasso_vars))

```

```{r message=FALSE, warning=FALSE}

# Get ordered features from CART
ordered_features_from_importance <- var_importance_df %>%
  arrange(desc(importance)) %>%
  pull(variable)

# All available features
all_numeric <- test_results %>% 
  select(-ptid_e, -predicted, -pred_prob, -outcome, -correct) %>%
  select(where(is.numeric)) %>%
  names()

all_categorical <- test_results %>% 
  select(-ptid_e, -predicted, -pred_prob, -outcome, -correct) %>%
  select(where(~ is.character(.) || is.factor(.))) %>%
  names()

# Prioritized + fallback sort
numeric_features <- c(
  intersect(ordered_features_from_importance, all_numeric),
  setdiff(sort(all_numeric), ordered_features_from_importance)
)

categorical_features <- c(
  intersect(ordered_features_from_importance, all_categorical),
  setdiff(sort(all_categorical), ordered_features_from_importance)
)

```

```{r message=FALSE, warning=FALSE}

library(dplyr)
library(tidyr)
library(purrr)

# Prepare reference data 
test_results <- test_results %>%
  mutate(correct = ifelse(predicted == outcome, "Correct", "Misclassified"))

# Numeric reference stats
numeric_stats <- test_results %>%
  filter(correct == "Correct") %>%
  summarise(across(all_of(numeric_features), 
                   list(mean = ~ mean(., na.rm = TRUE)), 
                   .names = "{.col}"))


# T-test p-values for numeric features
t_test_summary <- list()

for (feature in numeric_features) {
  correct_vals <- test_results %>% filter(correct == "Correct") %>% pull(!!feature)
  misclassified_vals <- test_results %>% filter(correct == "Misclassified") %>% pull(!!feature)
  
  if (length(unique(na.omit(correct_vals))) > 1 && length(unique(na.omit(misclassified_vals))) > 1) {
    test_result <- tryCatch(
      t.test(correct_vals, misclassified_vals),
      error = function(e) NULL
    )
    
    if (!is.null(test_result)) {
      t_test_summary[[feature]] <- list(
        p_value = test_result$p.value,
        statistic = test_result$statistic,
        df = test_result$parameter
      )
    }
  }
}


# Chi-squared test p-values for categorical features
chi_squared_summary <- list()

for (feature in categorical_features) {
  tbl <- table(test_results[[feature]], test_results$correct)

  if (nrow(tbl) > 1 && ncol(tbl) > 1) {
    test_result <- tryCatch(
      chisq.test(tbl),
      error = function(e) NULL
    )

    if (!is.null(test_result)) {
      chi_squared_summary[[feature]] <- list(
        p_value = test_result$p.value,
        statistic = test_result$statistic,
        df = test_result$parameter
      )
    }
  }
}


# Categorical frequency deviations
categorical_stats <- test_results %>%
  pivot_longer(all_of(categorical_features), names_to = "feature") %>%
  count(feature, value, correct) %>%
  group_by(feature, correct) %>%
  mutate(total = sum(n), freq = n/total) %>%
  select(-n, -total) %>%
  pivot_wider(names_from = correct, values_from = freq) %>%
  mutate(
    deviation = case_when(
      is.na(Misclassified) ~ -1,
      is.na(Correct) ~ 1,
      TRUE ~ (Misclassified / pmax(Correct, 0.001)) - 1  # Prevent division by zero
    )
  ) %>%
  select(feature, category = value, deviation)



# Get feature-level p-values
## Numeric features
t_test_pvals <- map_dbl(numeric_features, ~ pluck(t_test_summary, .x, "p_value", .default = NA_real_)) %>%
  set_names(numeric_features)

## Categorical features
chi_squared_pvals <- map_dbl(categorical_features, ~ pluck(chi_squared_summary, .x, "p_value", .default = NA_real_)) %>%
  set_names(categorical_features)

# Build enriched table
process_features <- function(data) {
  map_dfc(c(numeric_features, categorical_features), function(feat) {
    # Get normalized CART importance
    cart_importance <- var_importance_df %>%
      filter(variable == feat) %>%
      pull(importance) %>%
      first() %||% NA_real_

    if(feat %in% numeric_features) {
      # Same checks
      if(var(data[[feat]], na.rm = TRUE) == 0) {
        pval_label <- "zero_variance"
      } 
      else if(max(table(data[[feat]]))/nrow(data) > 0.95) {
        pval_label <- "near_zero_variance"
      } 
      else {
        qnt <- quantile(data[[feat]], c(0.25, 0.75), na.rm = TRUE)
        iqr <- IQR(data[[feat]], na.rm = TRUE)
        if(any(data[[feat]] < (qnt[1] - 1.5*iqr) | data[[feat]] > (qnt[2] + 1.5*iqr))) {
          pval_label <- "outliers_detected"
        } else {
          pval_label <- pluck(t_test_summary, feat, "p_value", .default = NA_real_)
        }
      }

      tibble(
        !!paste0(feat, "_cart_importance") := cart_importance,
        !!paste0(feat, "_value") := data[[feat]],
        !!paste0(feat, "_deviation") := data[[feat]] - numeric_stats[[feat]],
        !!paste0(feat, "_pvalue") := pval_label
        
      )

    } else {
      categories <- data[[feat]]
      deviations <- map_dbl(categories, ~ {
        categorical_stats %>%
          filter(feature == feat, category == .x) %>%
          pull(deviation) %>%
          coalesce(0)
      })

      pval_label <- pluck(chi_squared_summary, feat, "p_value", .default = NA_real_)

      tibble(
        !!paste0(feat, "_cart_importance") := cart_importance,
        !!paste0(feat, "_value") := categories,
        !!paste0(feat, "_deviation") := deviations,
        !!paste0(feat, "_pvalue") := pval_label
      )
    }
  })
}

# Final enriched data
enriched_misclassified <- misclassified %>%
  select(ptid_e) %>%
  bind_cols(process_features(misclassified)) %>%
  bind_cols(misclassified %>% select(predicted, pred_prob, outcome))

# Save outputs
write_csv(enriched_misclassified, "reports/enriched_misclassified.csv")


```

Here is a summary for what each metric in below 'Misclassified Cohort Cases' table represents:

For each feature (e.g., bmi_at_transplant, serum_albumin_at_transplant, etc.), the table includes:\
- **CART Feature Importance**: The feature importance score from the CART model, indicating how much this feature contributes to the model's predictions. (Columns are ordered by feature importance)\
- **Value**: The actual value of the feature for the misclassified case from the test dataset\
- **Deviation**: The difference between the feature value for the misclassified case and the mean value of that feature among correctly classified cases\
- **P-value**: The p-value from statistical tests comparing the feature's distribution between correct and misclassified cases

The table also includes:\
- **ptid_e**: Patient Identifier\
- **predicted**: The model's prediction (0 or 1)\
- **pred_prob**: The model's predicted probability\
- **outcome**: The actual outcome (0 or 1)

**Interpretation**:\
- Large deviations suggest that extreme or atypical values of a feature are associated with misclassification\
- Low p-values indicate statistically significant differences in feature values between correct and misclassified cases\
- Higher CART importance values indicate features that are more influential in the model's decision-making\
- The combination of these metrics helps identify which features and values are most associated with model errors

### Misclassifed Cohort Table

```{r}

# Convert to interactive DT table
DT::datatable(
  enriched_misclassified,
  filter = 'top',
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    autoWidth = TRUE,
    columnDefs = list(
      list(targets = '_all', className = 'dt-center'),
      list(targets = c(0), width = '100px'),  # ptid_e
      list(targets = c(1:ncol(enriched_misclassified)-3), width = '150px')  # feature columns
    )
  ),
  caption = 'Misclassified Cohort Cases with Deviation Analysis',
  class = 'cell-border stripe'
) %>%
  DT::formatRound(columns = c('pred_prob'), digits = 3) %>%
  DT::formatStyle(
    'predicted',
    target = 'row',
    backgroundColor = DT::styleEqual(c(0, 1), c('#ffebee', '#e8f5e9'))
  )

```

```{r echo=FALSE}

library(dplyr)
library(tidyr)
library(pheatmap)
library(stringr)


feature_type_table <- tibble(
  feature = c(numeric_features, categorical_features),
  feature_type = c(rep("Numeric", length(numeric_features)),
                   rep("Categorical", length(categorical_features)))
)

# Ensure consistent types for pivoting
enriched_misclassified <- enriched_misclassified %>%
  mutate(across(ends_with("_pvalue"), as.character))

# Build long format
long_enriched <- enriched_misclassified %>%
  pivot_longer(
    cols = matches("_value$"),
    names_to = "feature",
    names_pattern = "(.*)_value",
    values_to = "value",
    values_transform = as.character
  ) %>%
  left_join(
    enriched_misclassified %>%
      pivot_longer(
        cols = matches("_importance$"), 
        names_to = "feature_importance",
        names_pattern = "(.*)_importance",
        values_to = "importance"
      ),
    by = c("ptid_e", "feature" = "feature_importance", "predicted", "pred_prob", "outcome")
  ) %>%
  left_join(
    enriched_misclassified %>%
      pivot_longer(
        cols = matches("_cart_importance$"), 
        names_to = "feature_cart_importance",
        names_pattern = "(.*)_cart_importance",
        values_to = "cart_importance"
      ),
    by = c("ptid_e", "feature" = "feature_cart_importance", "predicted", "pred_prob", "outcome")
  ) %>%
  left_join(
    enriched_misclassified %>%
      pivot_longer(
        cols = matches("_deviation$"),  
        names_to = "feature_deviation",
        names_pattern = "(.*)_deviation",
        values_to = "deviation"
      ),
    by = c("ptid_e", "feature" = "feature_deviation", "predicted", "pred_prob", "outcome")  
  ) %>%
  left_join(
    enriched_misclassified %>%
      pivot_longer(
        cols = matches("_pvalue$"), 
        names_to = "feature_pvalue",
        names_pattern = "(.*)_pvalue",
        values_to = "pvalue"
      ),
    by = c("ptid_e", "feature" = "feature_pvalue", "predicted", "pred_prob", "outcome") 
  ) %>%
  select(ptid_e, feature, value, importance, cart_importance, deviation, pvalue, predicted, pred_prob, outcome) %>%
  left_join(feature_type_table, by = "feature")

```

### Missed Predictions Heatmap

| Color | Meaning |
|------------------------------------|------------------------------------|
| 🔴 Red | Very large deviation or strong signal (e.g., low p-value) — likely influencing the model heavily |
| 🔵 Blue | No signal — close to population norm for correctly classified patients |
| ⚪ Light/white | Mild deviation — mild influence |

#### Full Misclassified Dataset

```{r}

heatmap_data <- long_enriched %>%
  mutate(
    heatmap_value = case_when(
      feature_type == "Numeric" ~ as.numeric(deviation),
      feature_type == "Categorical" ~ as.numeric(deviation),
      TRUE ~ 0
    )
  ) %>%
  select(ptid_e, feature, heatmap_value) %>%
  tidyr::pivot_wider(names_from = feature, values_from = heatmap_value) %>%
  as.data.frame()

rownames(heatmap_data) <- heatmap_data$ptid_e
heatmap_data$ptid_e <- NULL

heatmap_matrix <- as.matrix(heatmap_data)
heatmap_matrix[is.na(heatmap_matrix)] <- 0

pheatmap::pheatmap(
  heatmap_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  show_rownames = FALSE,
  show_colnames = TRUE,
  main = "Risk-Driving Features in Misclassified Patients"
)


```

#### Top Missed Features

```{r echo=FALSE, warning=FALSE, message=FALSE}
#| fig-width: 10
#| fig-height: 8
#| out-width: "100%"
#| fig-align: "center"

# Compute feature scores: use mean absolute deviation or -log10(pvalue)
top_features <- long_enriched %>%
  mutate(
    score = case_when(
      feature_type == "Numeric" ~ abs(as.numeric(deviation)),
      feature_type == "Categorical" ~ -log10(as.numeric(pvalue)),
      TRUE ~ 0
    )
  ) %>%
  group_by(feature) %>%
  summarise(mean_score = mean(score, na.rm = TRUE)) %>%
  arrange(desc(mean_score)) %>%
  slice_head(n = 6)  # pick top 6 based on cluster visual inspection

# Keep only those features
long_top <- long_enriched %>%
  filter(feature %in% top_features$feature)

heatmap_data <- long_top %>%
  mutate(
    heatmap_value = case_when(
      feature_type == "Numeric" ~ as.numeric(deviation),
      feature_type == "Categorical" ~ as.numeric(deviation),
      TRUE ~ 0
    )
  ) %>%
  select(ptid_e, feature, heatmap_value) %>%
  tidyr::pivot_wider(names_from = feature, values_from = heatmap_value) %>%
  as.data.frame()

rownames(heatmap_data) <- heatmap_data$ptid_e
heatmap_data$ptid_e <- NULL

heatmap_matrix <- as.matrix(heatmap_data)
heatmap_matrix[is.na(heatmap_matrix)] <- 0

pheatmap::pheatmap(
  heatmap_matrix,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  show_rownames = TRUE,
  fontsize_row = 8,
  show_colnames = TRUE,
  main = "Top Risk-Driving Features in Misclassified Patients"
)

```
